Using Dependent Types to Build Safe and Concise JavaScript Foreign Function Interfaces

### 2013-06-07 papers/gpce2013/gpce2013.tex
eod

\begin{abstract}
Writing large Web applications is known to be difficult. One challenge comes from the fact that the application's
logic is scattered into heterogeneous clients and servers, making it difficult to share code between both sides or to
move code from one side to the other. Another challenge is performance: while Web applications rely on ever more
code on the client-side, they may run on smart phones with little hardware capabilities. These two challenges raise
the following problem: how to benefit from high-level languages and libraries making code complexity easier to manage
and abstracting over the clients and servers differences without trading this engineering comfort for performance?
This article presents high-level abstractions defined as deep embedded DSLs in Scala, that can (1) generate efficient
code leveraging the target platform characteristics, (2) be shared between client and server code. Code written with
our DSL has a high level of abstraction but runs as fast as hand tuned low-level JavaScript code.
\end{abstract}

@@

Kossakowski \etal introduced \emph{js-scala}, a compiled embedded DSL defined in Scala that generates JavaScript
code, making it possible to write the client-side code of Web applications using
JavaScript~\cite{Kossakowski12_JsDESL}. However, Kossakowski \etal did not address the engineering dilemma described
above. This paper enriches \emph{js-scala} with the following contributions:

\begin{itemize}
 \item we define high-level abstractions typically used in Web programming, that generate low-level code leveraging
native Web browsers APIs;
 \item in the case of abstractions shared between servers and clients, we specialize the code generation in order to
leverage the target platform native environments.
\end{itemize}

@@

The remainder of this paper is organized as follows. The next section introduces existing approaches for defining
cross-compiling languages. Sections \ref{contribution} and \ref{contribution-shared} present our contribution.
Section \ref{implementation} gives implementation details. Section \ref{validation} evaluates the contribution.
Section \ref{discussion} concludes.

@@

Alternatively, one can define concepts relative to a specific domain as libraries on top of a thin general purpose
language (this is also referred to as \emph{embedded} domain specific languages). Figure \ref{langs} (c) depicts this
approach. The general purpose language is used as a host language and does not need to be modified if a new concept
is introduced, because concepts are defined as pure libraries. However, this approach gives no opportunity to
translate a concept efficiently according to the target platform characteristics. Examples of languages following
this approach are Java/GWT, Kotlin, HaXe and SharpKit.

@@

@@

In a Web application, the user interface is defined by a HTML document that can be updated by the JavaScript code.
A typical operation consists in searching some “interesting” element in the document, in order to extract its
content, replace it or listen to user events triggered on it (such as a mouse click). The standard API provides
several functions to search elements in a HTML document according to their name or attribute values.
Figure~\ref{selectors-api} summarizes the available functions and their differences.

@@

The \code{querySelector} and \code{querySelectorAll} are the most general functions while the others handle special
cases. For the developer it is not convenient to have to master several functions performing similar tasks. In fact,
most JavaScript developers use the jQuery
library~\cite{Bibeault08_jQuery}\footnote{According to the following document:
\href{http://trends.builtwith.com/javascript}{ http://trends.builtwith.com/javascript}, jQuery is used by more than
40\% of the top million sites} that provides only one high-level function to search for elements.
Listings \ref{vanilla-selectors} and \ref{jquery-selectors} show two equivalent JavaScript programs performing
element searches, the first one using the native APIs and the second one using jQuery.

\begin{figure}
\begin{lstlisting}[language=JavaScript,label=vanilla-selectors,caption=Selectors in plain JavaScript]
function getWords() {
  var form = document.getElementById('add-user');
  var sections =
    form.getElementsByTagName('fieldset');
  var results = [];
  for (var i = 0 ; i < sections.length ; i++) {
    var words = sections[i]
      .getElementsByClassName('word');
    results[i] = words;
  }
  return results
}
\end{lstlisting}
\end{figure}

\begin{figure}
\begin{lstlisting}[language=JavaScript,label=jquery-selectors,caption=Selectors in jQuery]
function getWords() {
  var form = $('#add-user');
  var sections = $('fieldset', form);
  return sections.map(function () {
    return $('.word', this)
  })
}
\end{lstlisting}
\end{figure}

@@

\begin{figure}
\begin{lstlisting}[label=js-scala-selectors,caption=Selectors in js-scala]
def getWords() = {
  val form = document.find("#add-user")
  val sections = form.findAll("fieldset")
  sections map (_.findAll(".word"))
}
\end{lstlisting}
\end{figure}

Listing~\ref{js-scala-selectors} shows how to implement listing~\ref{jquery-selectors} using \emph{js-scala}. We
provide two functions for searching elements: \code{find} to find the first element matching a selector and
\code{findAll} to find all the matching elements. During staging these functions try to analyze the selector that is
passed as parameter and, when appropriate, produce code using the specialized API, otherwise they produce code using
\code{querySelector} and \code{querySelectorAll}. As a result, listing~\ref{js-scala-selectors} generates a
JavaScript program identical to listing~\ref{vanilla-selectors}: the high-level abstraction (the \code{find} and
\code{findAll} functions) exist only in the initial program, not in the final JavaScript program. It gives two
advantages: (1) the execution of the final program is more performant because of the use of specialized APIs, (2) the
final program’s size is smaller because it does not need to include jQuery. (Move this last sentence in the
discussion)

@@

Such a monadic API brings both safety and expressiveness to developers manipulating optional values but usually
involves the creation of an extra container object holding the optional value. In our case, we expose
the monadic API at the first stage but we generate code that does not wrap values in a container object but instead
checks if they are \code{null} or not when dereferenced. So the extra container object exists only at the
stage-level and is removed during code generation: listing \ref{null-js-scala} produces a code equivalent to listing
\ref{null-defensive}.

@@

A common task in Web applications consists in computing HTML fragments representing a part of the page content. This
task can be performed either from the server-side (to initially respond to a request) or from the client-side (to
update the current page). As an example, listing \ref{dom-api} defines a JavaScript function \code{articleUi} that
builds a DOM tree containing an article description, and listing \ref{scala-xml-api} shows how one could implement
a similar function on server-side using the standard Scala XML library. The reader may notice that the client-side
and server-side API are very different and that the client-side API is very low-level and inconvenient to use.

@@

The \code{el} function returns an \code{Element} IR node that is a tree composed of other \code{Element} nodes and
\code{Text} nodes. This tree is traversed by the code generators to produce code building an equivalent DOM tree on
client-side and code building an equivalent XML fragment on server-side. When the children elements of an element are
constant values (as in listing \ref{forest}) rather than dynamically obtained (as in listing \ref{forest-loops}), the
code generators inline the loop that adds children to their parent, for better performance. As a result, listing
\ref{forest} generates a code equivalent to listing \ref{dom-api} on client-side and equivalent to
\ref{scala-xml-api} on server-side.

@@

\section{Implementation}
\label{implementation}

\footnote{The code is available at \href{http://github.com/js-scala}{http://github.com/js-scala}}

\subsection{Selectors}

\begin{figure}
\begin{lstlisting}[label=selector-impl,caption=Selectors optimization]
def find(receiver: Exp[Selector],
         selector: Exp[String]) =
  getConstIdentifier(selector) match {
    case Some(id) if receiver == document =>
      DocumentGetElementById(Const(id))
    case _ =>
      SelectorFind(receiver, selector)
  }
\end{lstlisting}
\end{figure}

Listing \ref{selector-impl}

\subsection{Null references}

Code generation consists in traversing the statement nodes produced by the program evaluation according to their
dependencies and to emit the code corresponding to each statement. LMS already sorts the statements graph so DSL
authors just need to say how to emit code for each statement node of their DSL. Listing \ref{option-codegen} shows
such a code generator for the \code{null} reference handling DSL. The \code{emitNode} method handles
\code{OptionIsEmpty} and \code{OptionForeach} nodes. In the case of the \code{OptionIsEmpty} node, it simply
generates an expression testing if the value is \code{null}, in the case of the \code{OptionForeach} node, it wraps
the code block dereferencing the value within a \code{if} checking that the value is not \code{null}.

\begin{figure}
\begin{lstlisting}[caption=Null reference handling DSL code generator,label=option-codegen]
trait JSGenOptionOps extends JSGenEffect {
  val IR: OptionOpsExp
  import IR._
  override def emitNode(sym: Sym[Any], rhs: Def[Any]) =
    rhs match {
      case OptionIsEmpty(o) =>
        emitValDef(sym, q" $o === null")
      case OptionForeach(o, a, b) =>
        stream.println(q"if ($o !== null) {")
        emitValDef(a, quote(o))
        emitBlock(b)
        stream.println("}")
      case _ =>
        super.emitNode(sym, rhs)
    }
}
\end{lstlisting}
\end{figure}

\subsection{DOM Fragments}

@@

We implemented several applications using js-scala. We also have written several implementations of a complete
application, using different approaches to write the client and server sides, and compared the amount of code
written, the runtime performances and the ability to modularize the code and to maintain it.

>>>>

\begin{abstract}
Writing large Web applications is known to be difficult. One challenge comes from the fact that the application's
logic is scattered into heterogeneous clients and servers, making it difficult to share code between both sides or to
move code from one side to the other. Another challenge is performance: while Web applications rely on ever more
code on the client-side, they may run on smart phones with little hardware capabilities. These two challenges raise
the following problem: how to benefit from high-level languages and libraries making code complexity easier to manage
and abstracting over the clients and servers differences without trading this engineering comfort for performance?
This article presents high-level abstractions defined as deep embedded DSLs in Scala, that can (1) generate efficient
code leveraging the target platform characteristics, (2) be shared between clients and servers. Code written with
our DSL has a high level of abstraction but runs as fast as hand tuned low-level JavaScript code.
\end{abstract}

@@

Kossakowski \etal introduced \emph{js-scala}, a compiled embedded DSL defined in Scala that generates JavaScript
code, making it possible to write the client-side code of Web applications using
JavaScript~\cite{Kossakowski12_JsDESL}. However, Kossakowski \etal did not address the engineering dilemma described
above. This paper enriches \emph{js-scala} with the following contributions\footnote{The code is available at
\href{http://github.com/js-scala}{http://github.com/js-scala}}:

\begin{itemize}
 \item we define high-level abstractions, typically used in Web programming, that generate low-level code leveraging
native Web browsers APIs;
 \item in the case of abstractions shared between servers and clients, we specialize the code generation in order to
leverage the target platform native environments.
\end{itemize}

@@

The remainder of this paper is organized as follows. The next section introduces existing approaches for defining
cross-compiling languages. Sections \ref{contribution} and \ref{contribution-shared} present our contribution.
Section \ref{validation} compares our solution to common approaches. Section \ref{discussion} concludes.

@@

Alternatively, one can define concepts relative to a specific domain as libraries on top of a thin general purpose
language (this is also referred to as \emph{embedded} domain-specific languages). Figure \ref{langs} (c) depicts this
approach. The general purpose language is used as a host language and does not need to be modified if a new concept
is introduced, because concepts are defined as pure libraries. However, this approach gives no opportunity to
translate a concept efficiently according to the target platform characteristics. Examples of languages following
this approach are Java/GWT, Kotlin, HaXe and SharpKit.

@@

Code generation consists in traversing the statement nodes produced by the program evaluation according to their
dependencies and to emit the code corresponding to each statement. LMS already sorts the statements graph so DSL
authors just need to say how to emit code for each statement node of their DSL.

@@

In a Web application, the user interface is defined by a HTML document that can be updated by the JavaScript code.
A typical operation consists in searching some “interesting” element in the document, in order to extract its
content, replace it or listen to user events triggered on it (such as mouse clicks). The standard API provides
several functions to search elements in a HTML document according to their name or attribute values.
Figure~\ref{selectors-api} summarizes the available functions and their differences.

@@

The \code{querySelector} and \code{querySelectorAll} are the most general functions while the others handle special
cases. For the developer it is not convenient to have to master several functions performing similar tasks. In fact,
most JavaScript developers use the jQuery
library~\cite{Bibeault08_jQuery}\footnote{According to \href{http://trends.builtwith.com/javascript}{
http://trends.builtwith.com/javascript}, jQuery is used by more than
40\% of the top million sites} that provides only one high-level function to search for elements.
Listings \ref{vanilla-selectors} and \ref{jquery-selectors} show two equivalent JavaScript programs performing
element searches, the first one using the native APIs and the second one using jQuery.

\begin{figure}
\begin{lstlisting}[language=JavaScript,label=vanilla-selectors,caption=Searching elements in plain JavaScript]
function getWords() {
  var form = document.getElementById('add-user');
  var sections =
    form.getElementsByTagName('fieldset');
  var results = [];
  for (var i = 0 ; i < sections.length ; i++) {
    var words = sections[i]
      .getElementsByClassName('word');
    results[i] = words;
  }
  return results
}
\end{lstlisting}
\end{figure}

\begin{figure}
\begin{lstlisting}[language=JavaScript,label=jquery-selectors,caption=Searching elements in jQuery]
function getWords() {
  var form = $('#add-user');
  var sections = $('fieldset', form);
  return sections.map(function () {
    return $('.word', this)
  })
}
\end{lstlisting}
\end{figure}

@@

\begin{figure}
\begin{lstlisting}[label=js-scala-selectors,caption=Searching elements in js-scala]
def getWords() = {
  val form = document.find("#add-user")
  val sections = form.findAll("fieldset")
  sections map (_.findAll(".word"))
}
\end{lstlisting}
\end{figure}

Listing~\ref{js-scala-selectors} shows how to implement listing~\ref{jquery-selectors} using \emph{js-scala}. We
provide two functions for searching elements: \code{find} to find the first element matching a selector and
\code{findAll} to find all the matching elements. During staging these functions try to analyze the selector that is
passed as parameter and, when appropriate, produce code using the specialized API, otherwise they produce code using
\code{querySelector} and \code{querySelectorAll}. As a result, listing~\ref{js-scala-selectors} generates a
JavaScript program identical to listing~\ref{vanilla-selectors}: the high-level abstraction (the \code{find} and
\code{findAll} functions) exist only in the initial program, not in the final JavaScript program. It gives two
advantages: (1) the execution of the final program is more performant because of the use of specialized APIs, (2) the
final program’s size is smaller because it does not need to include jQuery. (Todo. Move this last sentence in the
discussion)

\begin{figure}
\begin{lstlisting}[label=selector-impl,caption=Selectors optimization]
def find(receiver: Exp[Selector],
         selector: Exp[String]) =
  getConstIdCss(selector) match {
    case Some(id) if receiver == document =>
      DocumentGetElementById(Const(id))
    case _ =>
      SelectorFind(receiver, selector)
  }
\end{lstlisting}
\end{figure}

Listing \ref{selector-impl} shows the implementation of the \code{find} function producing different IR node
according to the selector passed as parameter. The \code{getConstIdCss} function analyzes the selector: if it
is a constant \code{String} value containing a CSS ID selector, it returns the value of the identifier. If the
\code{find} function is applied to the \code{document} and to an ID selector, it returns a
\code{DocumentGetElementById} IR node (that is translated to a \code{document.getElementById} call by the code
generator), otherwise it returns a \code{SelectorFind} IR node (that is translated to a \code{querySelector}
call). (FIXME Add a ref to the spec of CSS selectors?)

The same applies to the implementation of \code{findAll}: the selector passed as parameter is analyzed and the
function returns a \code{SelectorGetElementsByClassName} in case of a CSS class name selector, a
\code{SelectorGetElementsByTagName} in case of a CSS tag name selector, and a \code{SelectorFindAll} otherwise.

@@

Such a monadic API brings both safety and expressiveness to developers manipulating optional values but usually
involves the creation of an extra container object holding the optional value. In our case, we expose the monadic API
at the first stage but we generate code that does not wrap values in container objects but instead checks if they are
\code{null} or not when dereferenced. So the extra container object exists only at the stage-level and is removed
during code generation: listing \ref{null-js-scala} produces a code equivalent to listing \ref{null-defensive}.

\begin{figure}
\begin{lstlisting}[caption=JavaScript code generator for null references handling DSL,label=option-codegen]
override def emitNode(sym: Sym[Any], rhs: Def[Any]) =
  rhs match {
    case OptionIsEmpty(o) =>
      emitValDef(sym, q" $o === null")
    case OptionForeach(o, b) =>
      stream.println(q"if ($o !== null) {")
      emitBlock(b)
      stream.println("}")
    case _ =>
      super.emitNode(sym, rhs)
  }
\end{lstlisting}
\end{figure}

Listing \ref{option-codegen} shows the JavaScript code generator for methods \code{isEmpty} (that checks if the
optional value contains a value) and \code{foreach} (that is called when the \code{for} notation is used, as in
listing \ref{null-js-scala}). The \code{emitNode} method handles \code{OptionIsEmpty} and \code{OptionForeach} nodes
returned by the implementations of \code{isEmpty} and \code{foreach}, respectively. In the case of the
\code{OptionIsEmpty} node, it simply generates an expression testing if the value is \code{null}. In the case of the
\code{OptionForeach} node, it wraps the code block dereferencing the value within a \code{if} checking that the value
is not \code{null}.

@@

A common task in Web applications consists in computing HTML fragments representing a part of the page content. This
task can be performed either from the server-side (to initially respond to a request) or from the client-side (to
update the current page). As an example, listing \ref{dom-api} defines a JavaScript function \code{articleUi} that
builds a DOM tree containing an article description, and listing \ref{scala-xml-api} shows how one could implement
a similar function on server-side using the standard Scala XML library. The reader may notice that the client-side
and server-side APIs are very different and that the client-side API is very low-level and inconvenient to use.

@@

The \code{el} function returns an \code{Element} IR node that is a tree composed of other \code{Element} nodes and
\code{Text} nodes. This tree is traversed by the code generators to produce code building an equivalent DOM tree on
client-side and code building an equivalent XML fragment on server-side. When the children of an element are constant
values (as in listing \ref{forest}) rather than dynamically computed (as in listing \ref{forest-loops}), the code
generators inline the loop that adds children to their parent, for better performance. As a result, listing
\ref{forest} generates a code equivalent to listing \ref{dom-api} on client-side and equivalent to
\ref{scala-xml-api} on server-side.

\begin{figure}
\begin{lstlisting}[language=JavaScript,label=js-gen-forest,caption=JavaScript code generator for the DOM fragment
definition DSL]
case Tag(name, children, attrs) =>
  emitValDef(sym, q"document.createElement('$name')")
  for ((n, v) <- attrs) {
    stream.println(q"$sym.setAttribute('$n', $v);")
  }
  children match {
    case Left(children) =>
      for (child <- children) {
        stream.println(q"$sym.appendChild($child);")
      }
    case Right(children) =>
      val x = fresh[Int]
      stream.println(q"for (var $x = 0; $x < $children.length; $x++) {")
      stream.println(q"$sym.appendChild($children[$x]);")
      stream.println("}")
  }
case Text(content) =>
  emitValDef(sym, q"document.createTextNode($content)")
\end{lstlisting}
\end{figure}

\begin{figure}
\begin{lstlisting}[label=scala-gen-forest,caption=Scala code generator for the DOM fragment definition DSL]
case Tag(name, children, attrs) =>
  val attrsFormatted =
    (for ((name, value) <- attrs)
     yield q" $name={ $value }").mkString
  children match {
    case Left(children) =>
      if (children.isEmpty) {
        emitValDef(sym, q"<$name$attrsFormatted />")
      } else {
        emitValDef(sym,
          q"<name$attrsFormatted>{ ${children.map(quote)} }</$name>"
        )
      }
    case Right(children) =>
      emitValDef(sym, q"<$name$attrsFormatted>{ $children }</$name>")
  }
case Text(content) =>
  emitValDef(sym, q"{xml.Text(content)}")
\end{lstlisting}
\end{figure}

Listings \ref{js-gen-forest} and \ref{scala-gen-forest} show the relevant parts of the code generators for this DSL.
They basically follow the same pattern: they visit \code{Tag} and \code{Text} IR nodes and produce the corresponding
elements in the target language. (TODO More details)

@@

@@

We measured the runtime performances of the DSL for optional values using a micro-benchmark.

We also have written several implementations of a complete application using different approaches for the client and
server sides, and compared the amount of code written, the runtime performances and the ability to modularize the
code and to maintain it.

\subsection{Micro-Benchmark}

### 2013-06-07 papers/sle2013/sle2013.tex
eod

\title{FFI Is Not Enough. We Need Dependent Types.}
\titlerunning{Dependent Types Rock}

@@

\begin{abstract}
Several programming languages can target JavaScript as a back-end, giving developers programming language features
that are absent from JavaScript, such as static typing. However, the Web browser APIs, which are needed to interact
with a Web page, are designed for JavaScript, making it challenging to expose them in a statically typed language.
Indeed, existing statically typed languages exposing Web browser APIs either break type safety or give developers
less control than if they were using JavaScript. This article shows how to expose Web browser APIs in Scala in a
type safe way while keeping the same level of control as with native APIs. We use advanced type system features such
as dependent types and functional dependencies.
\keywords{Dependent Types}
\end{abstract}

>>>>

\title{Using Dependent Types to Build Safe and Concise JavaScript Foreign Function Interfaces}
\titlerunning{Dependent Types Rock}

@@

\begin{abstract}
Several programming languages can target JavaScript as a back-end, giving developers programming language features
that are absent from JavaScript, such as static typing. However, the Web browser APIs, which are needed to interact
with a Web page, are designed for JavaScript, making it challenging to expose them in a statically typed language.
Indeed, existing statically typed languages exposing Web browser APIs either break type safety or give developers
less control than if they were using JavaScript. This article shows how to expose Web browser APIs in Scala in a
type safe way while keeping the same level of control as with native APIs by using dependent types and functional
dependencies. We validate this approach in designing safe and concise foreign function interfaces between Scala and
JavaScript for DOM events handling and DOM manipulation. We compared this approach to common frameworks such as GWT,
Fay, Kotlin, Dart and SharpKit.
\keywords{Dependent Types, JavaScript, Scala, Foreign Function Interface}
\end{abstract}
